\vspace{10em}

\setlength{\parskip}{1em}

During the years of my doctoral studies I had the luck to meet a number of great people, without whom this Thesis would never have been created. I would like to thank every one of them for their support, inspiration and friendship.

%
% Promotorzy
%

I would like to express my deepest gratitude to prof. Pawe\l{} Moskal for the opportunity to work in his research group, for his supervision over the preparation of this Thesis, and --- above all --- for his contagious passion for science.

The second person without whom this work would not be possible is dr Eryk Czerwi\'nski. I~am greatly indebted to Eryk for his guidance in all my research for the past six years and, perhaps even more importantly, for introducing me to virtually all aspects of academic life.
%
% Antonio
%

I would like to extend my special thanks to prof. Antonio Di Domenico, for giving me the possibility to work on the fascinating subject of direct symmetry tests with neutral kaons at KLOE and KLOE-2, and for his careful supervision on my analysis.

%
% Kamys i Jarczyk
%

I am grateful to prof.\ Bogus\l{}aw Kamys for the opportunity to work in the Department of Nuclear Physics of the Jagiellonian University and to prof.\ Lucjan Jarczyk for all his comments on my work, always motivating me to improve my research and presentation skills.

% 
% Koledzy z KLOE w Krakowie: Michal, Wojtek, Daria, Krzysiek
%

The time of my work in the Krak\'ow subgroup of KLOE was exceptional thanks my Colleagues Daria Kisielewska, Krzysztof Kacprzak, dr Wojciech Krzemie\'n and dr Micha\l{} Silarski. Thank you for the great and inspiring atmosphere and countless help you gave me during these years. 

%
% KLOE advice
%

I would like to thank prof.\ Filippo Ceradini, dr~Erika De~Lucia, dr~Antonio De~Santis, dr~Paolo Gauzzi and dr~Enrico Graziani for sharing their expertise in working with the KLOE data and for their suggestions which helped me overcome several dead ends in my work. I am also indebted to prof. Wojciech Wiślicki for his helpful advice on statistics.
%
% KLOE Colleagues
%

Moreover, my frequent visits in the Laboratories of Frascati would not be the same without the great people I met there and their hospitality, especially dr Elena Perez del~Rio, dr~Marcin Ber\l{}owski, dr~Paolo Fermani, dr~Gianfranco Morello and all my Colleagues from KLOE.
%

No less do I owe to my Colleagues from J-PET with whom I have worked on the second part of this Thesis. I would like to especially thank dr Magdalena Skurzok, Monika Pawlik-Niedźwiecka, dr Grzegorz Korcyl, Szymon Niedźwiecki and dr Sushil Sharma for providing me with building blocks for the calibration and analysis of J-PET data based on their great efforts.
%

Great thanks also to my officemate Krzysztof Nowakowski, for motivating me when it was time to write, and for saving my sanity with conversations about everything but science when it was time to take a break.


%
% Spradzić czy "to owe big" jest formalne
% Krzysiek?

%
% Rodzina
%

As so many others things in my life, this Thesis would never come to life without the continuous support of my wife Kasia, whom I would like to thank for her patience and unwavering belief that I will finish writing some day. I also owe great thanks to my son Andrzej, for bringing a completely new quality to my life in my last PhD student years, and for his unstoppable will to help me in writing of this text, even if by typing randomly on daddy's keyboard.

\begin{otherlanguage}{polish}
Na koniec dziękuję Wam, Mamo i Tato, za wiarę we mnie i za całą pomoc w dotarciu do tego momentu.
\end{otherlanguage}

\vspace{3cm}

This work was supported by the Polish National Science Centre through\\Projects No.~2014/14/E/ST2/00262 and 2016/21/N/ST2/01727.
%
% Może o grancie?
% 

%
% TODO: spraedzic czy gauzzi i Graziani to doktorzy
%


%%% Local Variables:
%%% TeX-master: "main"
%%% End: 